---
title: "Analyses for Longevity Manuscript"
author: "Stefano Monti"
date: 6/18/2019
output:
  html_notebook:
    toc: yes
  html_document:
    theme: united
    toc: yes
---

```{r global, echo=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE,fig.path='../results/manuscript/figures/')
```

```{r settings}
require(pheatmap)
require(openxlsx)
require(Biobase)
require(limma)
require(cba)
require(VennDiagram)

## SCC 
if ( grepl("scc",system("hostname",intern=TRUE)) ) {
    ## NECS is /restricted/projectnb/necs/
    PATH <- file.path( Sys.getenv("NECS"),"novartis-data/smonti")
}
## MAC
if ( grepl("bumc.bu.edu",system("hostname",intern=TRUE)) || grepl("Mac",system("hostname",intern=TRUE)) ) {
    ##require(hypeR)
    PATH <- "~/research/projects/longevity/novartis"
}
source(file.path(PATH,"scripts/support.functions.R"))
source(file.path(Sys.getenv("CBMRtools"),"CBMRtools/R/misc.R"))
```

## To-Do List

1. Can we make a Venn diagram of 4 lists:

	a. Aging signature

	b. List of 50 markers linked to extreme old age signature

	c. Longevity signature in offspring

	d. Longevity signature in centenarians

2. We analyzed offspring _vs control but nothing passed the
   significance level after correction. What if we restrict the
   analysis to the list of biomarkers on the longevity signature in
   the offspring, or the combined offspring/centenarians longevity
   signature? Do we see anything significant in this smaller set?
 
3. Creae heatmap for the hallmark pathways in the offspring longevity
   signature restricted to the following pathways:
   ESTROGEN_RESPONSE_LATE, ANGIOGENESIS, MTORC1_SIGNALING,
   TGF_BETA_SIGNALING, INFLAMMATORY_RESPONSE, DNA_REPAIR,
   IL6_JAK_STAT3_SIGNALING, ALLOGRAFT_REJECTION, UV_RESPONSE_DN,
   P53_PATHWAY, ANDROGEN_RESPONSE, WNT_BETA_CATENIN_SIGNALING
 

```{r load.data}
soma2 <- readRDS(file.path(PATH,"workdir/somascan/soma2.RDS"))
```

### 1. Four-signature Overlap

```{r 4signature.overlap}
## aging signature 
lmRes <- readRDS(file=file.path(PATH,"workdir/somascan/lmRes.RDS"))
## longevity signatures
centRes2 <- readRDS(file=file.path(PATH,"workdir/manuscript/centRes2.RDS"))
restRes2 <- readRDS(file=file.path(PATH,"workdir/manuscript/restRes2.RDS"))
## curated signatures
longevitySig <- readRDS(file=file.path(PATH,"gdrive/longevitySignatures.RDS"))

sigList <- list(
    aging=unique(lmRes[lmRes$fdr<=0.01,"geneID"]),
    longevityCent=unique(centRes2[centRes2$p.value<=.01,"symbol"]),
    longevityOffs=unique(restRes2[restRes2$p.value<=.01,"symbol"]),
    longevityNECS=longevitySig$NECSunique)
print(sapply(sigList,length))

fill <- c("light blue", "pink","orange","yellow")
size  <- rep(0.5,4)
venn <- venn.diagram(x = sigList,
                     filename = NULL,
                     height = 1000,
                     width = 2000, fill = fill,
                     cat.default.pos = "text",
                     cat.cex = size,
                     main = "Overlap of Aging/Longevity signatures",
                     hyper.test=TRUE,lower.tail=FALSE,total.population=length(unique(fData(soma2)$symbol)))
#grid.newpage()
grid.draw(venn)
```

### 2. Offspring vs. Control

Even after restricting to top 'genes' (q<=0.25) from the offsprings'
longevity signature, no genes pass significance when comparing
offsprings to controls.

```{r offspring.vs.control}
## restrict to non-centenarians and to longevity signature proteins
off2 <- soma2[rownames(restRes2)[restRes2$FDR<=0.25],soma2$Cohort!="Centenarian"]
off2$Cohort <- droplevels(off2$Cohort)
table(off2$Cohort)

diffOff <- as.data.frame(
    t(sapply( 1:nrow(off2),
             function(i) {
                 tmp <- lm(exprs(off2)[i,] ~ Cohort + Gender + Year.completed, data=pData(off2))
                 summary(tmp)$coefficients["CohortOffspring",c("Estimate","Pr(>|t|)")]
             }))
)
rownames(diffOff) <- featureNames(off2)
diffOff <- cbind(geneID=fData(off2)$symbol,diffOff)
diffOff <- diffOff[order(diffOff[,"Pr(>|t|)"]),]
diffOff$fdr <- p.adjust(diffOff[,"Pr(>|t|)"], method="BH")

print( rbind(analytes=c("fdr<=0.25"=sum(diffOff$fdr<=.05), "fdr<=0.05"=sum(diffOff$fdr<=.01)),
             proteins=c("fdr<=0.25"=length(unique(diffOff$geneID[diffOff$fdr<=.05])),
                 "fdr<=0.05"=length(unique(diffOff$geneID[diffOff$fdr<=.01])))) )
```

### 3. Offsprings Heatmap

```{r offsprings.heatmap}
pheatmapColors <- colGradient(c("darkgreen","white","darkorange"),length=11)

gsets <- c("ESTROGEN_RESPONSE_LATE",
           "ANGIOGENESIS",
           "MTORC1_SIGNALING",
           "TGF_BETA_SIGNALING",
           "INFLAMMATORY_RESPONSE",
           "DNA_REPAIR",
           "IL6_JAK_STAT3_SIGNALING",
           "ALLOGRAFT_REJECTION",
           "UV_RESPONSE_DN",
           "P53_PATHWAY",
           "ANDROGEN_RESPONSE",
           "WNT_BETA_CATENIN_SIGNALING")

rest1 <- readRDS(file.path(PATH,"workdir/rest1.RDS"))
table(rest1$age.group,rest1$FU.2group,useNA="ifany") 
somaHall <- readRDS(file.path(PATH,"workdir/somascan/somaHall.RDS"))
featureNames(somaHall) <- gsub("HALLMARK_","",featureNames(somaHall))
somaHall <- somaHall[match(gsets,featureNames(somaHall)),sampleNames(rest1)]
pData(somaHall) <- pData(rest1)
somaHall <- somaHall[,order(somaHall$FU.2group)] 

annot_col <- pData(somaHall)[,c("Cohort","age.group","FU.2group")]
annotColors <- list(age.group=c("lightgray","darkgray"),
                    FU.2group=c("lightblue","pink"),
                    Cohort=c("yellow2","yellow3"),
                    markers=c(up="darkred",down="darkgreen"))
names(annotColors$age.group) <- levels(somaHall$age.group)
names(annotColors$FU.2group) <- levels(somaHall$FU.2group)
names(annotColors$Cohort) <- levels(somaHall$Cohort)

gspfile <- paste0("workdir/manuscript/somaHall.rest2.longevity.gspDiffanal.RDS")
gspDiff <- readRDS(file.path(PATH,gspfile))
 
annot_row <-
    data.frame(markers=c("down","up")[as.integer(gspDiff[gsets,"Estimate"]>0)+1])
rownames(annot_row) <- gsets
annotColors$markers <- c(down="darkgreen",up="darkred")

hcRow <- hcopt(as.dist(1-cor(t(exprs(somaHall)),use="pairwise.complete.obs")),method="ward.D")
hcCol <- hcopt(dist(t(exprs(somaHall))),,method="ward.D")

pheatmap(exprs(somaHall),main="samples ordered by phenotype, markers clustered",
         color=pheatmapColors,
         annotation_col = annot_col,
         annotation_row = annot_row,
         annotation_colors = annotColors,
         cluster_rows=hcRow,
         cluster_cols=FALSE, #hcCol,
         show_rownames = TRUE,
         show_colnames = FALSE,
         labels_row=gsub("HALLMARK_","",featureNames(somaHall)),
         fontsize_row=9,
         scale = "row",
         cellheight=12)

pheatmap.group(mat=somaHall,main="samples clustered within phenotype, markers clustered",
               group="FU.2group",
               color=pheatmapColors,
               annotation_col = annot_col,
               annotation_row = annot_row,
               annotation_colors = annotColors,
               cluster_rows=TRUE,
               show_rownames = TRUE,
               show_colnames = FALSE,
               labels_row=gsub("HALLMARK_","",featureNames(somaHall)),
               fontsize_row=9,
               scale = "row",
               cellheight=12)
```
