---
title: "Somascan GeneSet Projection Analysis"
author: "Stefano Monti, Paola Sebastiani, Anastasia Gurinovich"
output:
  html_notebook:
    toc: yes
  html_document:
    theme: united
    toc: yes
---

```{r global, echo=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE,fig.path='../results/somascan/figures/')
```

```{r settings}
require(CBMRtools)
require(limma)
require(pheatmap)
require(openxlsx)
require(GSVA)
PATH <- "~/research/projects/longevity"
source(file.path(PATH,"scripts/support.functions.R"))
source(file.path(Sys.getenv("CBMRtools"),"dvlp/diffanalOverlap.R"))
overwrite <- FALSE
```

## GeneSet Projection

We next project the data onto the space of the <a
href="http://software.broadinstitute.org/gsea/msigdb/collections.jsp#C2">`C2.CP`</a>
and the <a
href="http://software.broadinstitute.org/gsea/msigdb/collections.jsp#h.all">`hallmarks`</a>
geneset compendia.

```{r data.preprocess}
## Load the necessary data. Since GSVA doesn't handle missing data, we
## take the unthresholded versions of the data
soma0 <- readRDS(file.path(PATH,"workdir/somascan/soma0.RDS")) # unthresholded data (hence, w/o NA's)
soma2 <- readRDS(file.path(PATH,"workdir/somascan/soma2.RDS")) # thresholded data w/ simplified pData
if ( !all.equal(dimnames(soma0),dimnames(soma2)) ) stop( "dimnames(soma0)!=dimnames(soma2)" )

soma3.file <- file.path(PATH,"workdir/somascan/soma3.RDS")
soma3 <- {
    if ( file.exists(soma3.file) && !overwrite ) {
        readRDS(soma3.file)
    }
    else {
        soma3 <- soma0                # use the data w/o thresholding
        pData(soma3) <- pData(soma2)  # ..but use the simplified annotation
                                      # annotate rows w/ gene symbols (for gsva)
        fData(soma3)$orig.symbol <- toupper(fData(soma3)$orig.symbol)
        soma3 <- collapse2genesymbol(eset=soma3, colId="orig.symbol", method='max')
        ## fData(soma3)$symbol <- toupper(fData(soma3)$symbol)
        ## soma3 <- collapse2genesymbol(eset=soma3, colId="symbol", method='max')
        saveRDS(soma3,file=soma3.file)
        soma3
    }
}
## load pheatmap annotation from 'analysis01.somascan.Rmd'
tmp <- readRDS(file.path(PATH,"workdir/somascan/pheatmap.annot.RDS"))
annot <- tmp$annot
annotCol <- tmp$annotCol
annot <- annot[sampleNames(soma3),]

## keep only centenarian annotation plus 'longevity follow-up' (see
## .. Longevity Grouping in analaysis01.somascan.Rmd)
annot <- annot[,"group",drop=FALSE]
annot$FU.2group <- soma3$FU.2group
annotCol <- c(annotCol,list(FU.2group=c(short="pink",long="green")))

## modification on 3/2/2019 (remove fu group, and group into centenarian, offspring, control)
annot <- annot[,"group",drop=FALSE]
grp <- as.character(annot[,"group"])
grp[grep("Offspring",grp)] <- "offspring"
grp[grep("Control",grp)] <- "control"
grp[grep("Centenarian",grp)] <- "centenarian"
annot[,"group"] <- factor(grp)
annotCol$group <- c(centenarian="red",offspring="pink",control="antiquewhite")
```

Since there are only `r nrow(soma3)` unique mapped genes represented
in the Somascan assay, we reduce the genesets' size accordingly, by
taking the overlap of each original geneset with the Somascan genes,
and by further eliminating the genesets with less than 5 members
(`preprocess.geneset` is defined in the file `support.functions.R`).

```{r gsp.preprocess.gsets}
c2cp <- preprocess.gsets(gsetsFile="~/researchData/annot/c2.cp.v6.1.symbols.gmt",
                         bkg=featureNames(soma3), minGset=5, do.unique=FALSE )
hall <- preprocess.gsets(gsetsFile="~/researchData/annot/h.all.v6.1.symbols.gmt",
                         bkg=featureNames(soma3), minGset=5, do.unique=FALSE )
```

We thus reduced the final list of C2.CP genesets to `r length(c2cp)`
sets, while all hallmarks 'survive' the filtering.

On these lists, we run <a
href="https://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-14-7">`GSVA`</a>,
wich computes a sample-specific geneset enrichment score (between -1.0
and 1.0) for each of the provided genesets, and generates a
corresponding geneset-by-sample matrix.

### C2.CP projection

```{r gsva.c2cp}
#soma3 <- readRDS(file.path(PATH,"workdir/somascan/soma3.RDS"))
somaC2cp.file <- file.path(PATH,"workdir/somascan/somaC2cp.RDS")
if ( file.exists(somaC2cp.file) && !overwrite ) {
    somaC2cp <- readRDS(somaC2cp.file)
}
if ( !file.exists(somaC2cp.file) || overwrite ) {
    somaC2cp <- gsva(soma3, c2cp, mx.diff=FALSE, verbose=FALSE, parallel.sz=1)$es.obs
    saveRDS(somaC2cp,file=somaC2cp.file)
}
featureNames(somaC2cp) <- gsub("KEGG_","K_",featureNames(somaC2cp))
featureNames(somaC2cp) <- gsub("REACTOME_","R_",featureNames(somaC2cp))
featureNames(somaC2cp) <- gsub("PID_","P_",featureNames(somaC2cp))
featureNames(somaC2cp) <- gsub("BIOCARTA_","B_",featureNames(somaC2cp))

hcRow <- hcopt(dist(exprs(somaC2cp)),method="ward.D")
hcCol <- hcopt(dist(t(exprs(somaC2cp))),method="ward.D")
pheatmap(exprs(somaC2cp),
         color=colGradient(c("blue","white","red"),length=15),
         annotation_col = annot,
         annotation_colors = annotCol,
         cluster_rows=hcRow,
         cluster_cols=hcCol,
         show_rownames = TRUE,
         show_colnames = FALSE,
         fontsize_row=2,
         labels_row=sapply(featureNames(somaC2cp),substr,1,20),
         scale = "none")
write.xlsx(exprs(somaC2cp)[rev(hcRow$order),rev(hcCol$order)],rowNames=TRUE,
           file=file.path(PATH,"results/somascan/somaC2cp.xlsx"))
```
The GSVA-projected data in excel format (and sorted according to
clustering) is stored in <a
href="somascan/somaC2cp.xlsx">`results/somascan/somaC2cp.xlsx`</a>.

### Hallmarks projection

Let's repeat by using the <a
href="http://software.broadinstitute.org/gsea/msigdb/collections.jsp#h.all">`hallmarks`
compendium</a>.

```{r gsva.hallmarks}
somaHall.file <- file.path(PATH,"workdir/somascan/somaHall.RDS")
if ( file.exists(somaHall.file) && !overwrite ) {
    somaHall <- readRDS(somaHall.file)
}
if ( !file.exists(somaHall.file) || overwrite ) {
    somaHall <- gsva(soma3, hall, mx.diff=FALSE, verbose=FALSE, parallel.sz=1)$es.obs
    saveRDS(somaHall,file=somaHall.file)
}
featureNames(somaHall) <- gsub("HALLMARK_","",featureNames(somaHall))

hcRow <- hcopt(dist(exprs(somaHall)),method="ward.D")
hcCol <- hcopt(dist(t(exprs(somaHall))),method="ward.D")
pheatmap(exprs(somaHall),
         color=colGradient(c("blue","white","red"),length=15),
         annotation_col = annot,
         annotation_colors = annotCol,
         cluster_rows=hcRow,
         cluster_cols=hcCol,
         show_rownames = TRUE,
         show_colnames = FALSE,
         fontsize_row=5,
         labels_row=sapply(featureNames(somaHall),substr,1,20),
         scale = "none")
write.xlsx(exprs(somaHall)[rev(hcRow$order),rev(hcCol$order)],rowNames=TRUE,
           file=file.path(PATH,"results/somascan/somaHall.xlsx"))
```

The GSVA-projected data in excel format (and sorted according to
clustering) is stored in <a
href="somascan/somaHall.xlsx">`results/somascan/somaHall.xlsx`</a>.

## Differential Pathway Analysis

We now perform differential analysis in the space of geneset
projection scores, both for C2.CP and hallmarks.

```{r diffanal.gsp.c2}
## defining a function for repeated tasks (in scripts/support.functions.R)
print(gsets.diffanal)

## perform diffanal based on C2.CP
lmC2cp <- gsets.diffanal(gspDat.file=file.path(PATH,"workdir/somascan/somaC2cp.RDS"),
                         do.heatmap=TRUE, annot=annot, annotCol=annotCol, fontsize_row=3)

## ..and on Hallmarks
lmHall <- gsets.diffanal(gspDat.file=file.path(PATH,"workdir/somascan/somaHall.RDS"),max.fdr=0.05,
                         do.heatmap=TRUE, annot=annot, annotCol=annotCol,fontsize_row=6)
```

## APOE QTL

```{r apoe.qtl.support.functions, echo=FALSE}
eQTL.simple.fun <- function( eset )
{
    eQTL <- data.frame(t(apply( exprs(eset), 1, function(Y) {
        fit <- lm(Y ~ eset$apoe)
        anv <- anova(fit)
        c( fit$coefficients[c("eset$apoeE2","eset$apoeE4")],unlist(anv[1,c("F value","Pr(>F)")]) )
    })),check.names=FALSE)
    colnames(eQTL) <- gsub("eset\\$","",colnames(eQTL))
#    eQTL <- cbind(eQTL,symbol=fData(eset)[,"hgnc_symbol"])
    eQTL <- eQTL[,c("symbol","apoeE2","apoeE4","F value","Pr(>F)")]
    eQTL$fdr <- p.adjust(eQTL[,"Pr(>F)"],method="BH")
    eQTL
}
eQTL.fun <- function( eset )
{
    eQTL <- data.frame(t(apply( exprs(eset), 1, function(Y) {
        fit2 <- lm(Y ~ eset$Cohort + eset$apoe)
        fit1 <- lm(Y ~ eset$Cohort)
        av12 <- anova(fit1,fit2)
        c( fit2$coefficients[c("eset$apoeE2","eset$apoeE4")],unlist(av12[2,c("F","Pr(>F)")]) )
    })),check.names=FALSE)
    colnames(eQTL) <- gsub("eset\\$","",colnames(eQTL))
#    eQTL <- cbind(eQTL,symbol=fData(eset)[,"hgnc_symbol"])
    eQTL <- eQTL[,c("apoeE2","apoeE4","F","Pr(>F)")]
    eQTL$fdr <- p.adjust(eQTL[,"Pr(>F)"],method="BH")
    eQTL
}
```

```{r apoe.eqtl}
somaC2cp <- readRDS(file=file.path(PATH,"workdir/somascan/somaC2cp.RDS"))
somaHall <- readRDS(file=file.path(PATH,"workdir/somascan/somaHall.RDS"))
apoe <- factor(rep("NA",ncol(somaHall)),levels=c("E3","E2","E4"))
apoe[somaC2cp$APOE %in% c("e2e2","e2e3")] <- "E2"
apoe[somaC2cp$APOE %in% c("e3e3")] <- "E3"
apoe[somaC2cp$APOE %in% c("e3e4")] <- "E4"
table(apoe,useNA="ifany")

if ( !all.equal(pData(somaC2cp),pData(somaHall)) ) stop( "pData(somaC2cp) != pData(somaHall)" )
somaC2cp$apoe <- somaHall$apoe <- apoe
somaC2cp1 <- somaC2cp[,!is.na(somaC2cp$apoe)]
somaHall1 <- somaHall[,!is.na(somaHall$apoe)]

eQTL.hall <- eQTL.fun( eset=somaHall1 )
eQTL.hall.ord <- eQTL.hall[order(eQTL.hall[,"fdr"],decreasing=FALSE),]
eQTL.c2cp <- eQTL.fun( eset=somaC2cp1 )
eQTL.c2cp.ord <- eQTL.c2cp[order(eQTL.c2cp[,"fdr"],decreasing=FALSE),]

saveRDS(eQTL.hall,file=file.path(PATH,"workdir/somascan/eQTL.hall.RDS"))
saveRDS(eQTL.c2cp,file=file.path(PATH,"workdir/somascan/eQTL.c2cp.RDS"))
write.xlsx(list(hallmarks=eQTL.hall.ord,c2cp=eQTL.c2cp.ord),rowNames=TRUE,
           file=file.path(PATH,"results/somascan/apoe.eQTL.GSVA.xlsx"))
```
