---
title: "Somascan Data Analysis"
author: "Stefano Monti, Paola Sebastiani, Anastasia Gurinovich"
output:
  html_notebook:
    toc: yes
  html_document:
    theme: united
    toc: yes
---

```{r global, echo=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE,fig.path='../results/somascan/figures/')
```

```{r settings}
require(CBMRtools)
require(limma)
require(pheatmap)
require(openxlsx)
require(GSVA)
library(Biobase)
PATH <- "~/research/projects/longevity/novartis"
source(file.path(PATH,"scripts/support.functions.R"))
source(file.path(Sys.getenv("CBMRtools"),"dvlp/diffanalOverlap.R"))
```

## Read and Process Somascan Data

Notice that there are multiple protein IDs mapping to the same gene
symbol. These will all be included in the analysis.

## QC analysis

We generate some exploratory plots to look at data distribution and possible outliers.

```{r qc.analysis}
soma <- readRDS(file.path(PATH,"data/somascan/180308_CentCollabBU_HybNormPlateScaleMedNormCal.eset.RDS"))
table(pData(soma)$Cohort)
```

We first look at the relationship between location (median) and scale
(median absolute deviation) of the log2-transformed data, which
confirms the suitability of an additive (linear) model.

```{r qc.mva}
## distribution of raw data
hist(exprs(soma))

## first, show the relationship btw location and scale
somaL <- soma; exprs(somaL) <- log2(exprs(soma))
hist(exprs(somaL))
loc <- apply(exprs(somaL),1,median)
scl <- apply(exprs(somaL),1,mad)
plot(loc,scl,log="",xlab="location (median)",ylab="scale (mad)")
```

Boxplots of within-sample distributions (in log2 space) confirm the
appropriateness of a log-normal model.

```{r qc.boxplots}
## then, show the within-sample distribution
cohortPalette <- c("pink","darkred","red","whitesmoke","lightgray")
names(cohortPalette) <- levels(pData(soma)[,"Cohort"])
cohortCol <- cohortPalette[pData(soma)[,"Cohort"]]
names(cohortCol) <- sampleNames(soma)

ord <- order(pData(somaL)$Cohort)
boxplot(exprs(somaL)[,ord],las=2,pch="-",col=cohortCol[ord],show.names=FALSE)
legend("bottomleft",col=cohortPalette,legend=names(cohortPalette),pch=15,cex=.75)
```

A first look at the heatmap of _all_ data shows the dominance of very
extreme values, which justifies the filtering we perform below.

```{r pheatmap}
## color gradient for the expression levels (blue=down-regulated; white=neutral; red=up-regulated)
bwrPalette <- colGradient(c("blue","white","red"),length=9)

## sample color coding
annot <- pData(soma)[, c("Cohort", "Health.Dementia", "Health.Stroke","Health.Heart.Attack")]
colnames(annot) <- c("group", "dementia", "stroke", "heart")

annotCol <- list(
    group = cohortPalette,
    dementia = c(No="white",Yes="red"),
    stroke = c(No="white",Yes="red"),
    heart = c(No="white",Yes="red")
)
saveRDS( list(annot=annot,annotCol=annotCol), file=file.path(PATH,"workdir/somascan/pheatmap.annot.RDS") )
hcCol <- hcopt(dist(t(exprs(somaL))),method="ward.D")
hcRow <- hcopt(as.dist(1-cor(t(exprs(somaL)))),method="ward.D")
pheatmap(exprs(somaL),
         color=bwrPalette,
         annotation_col = annot,
         annotation_colors = annotCol,
         cluster_rows=hcRow,
         cluster_cols=hcCol,
         show_rownames = FALSE,
         show_colnames = FALSE,
         scale = "row")
```

(NOTE: for clustering, we used the wrapper `hcopt` (defined in
CBMRtools), which adds an <a
href="https://academic.oup.com/bioinformatics/article-lookup/doi/10.1093/bioinformatics/btg030">optimal
leaf-ordering</a> to the standard `hclust` procedure).

## PCA projection

The PCA projection identifies 2 outliers (mostly evident in the 1st
component), which we will eliminate from subsequent analyses.

```{r pca}
pca <- prcomp(exprs(somaL), scale = TRUE) ## perform PCA
summary(pca)$importance[,1:4]             ## show variance explained by each component

## let's plot the first two components
if ( any(rownames(pca$rotation)!=names(cohortCol)) ) stop( "rownames(pca$rotation)!=names(cohortCol)" )
plot(x=pca$rotation[,"PC1"],y=pca$rotation[,"PC2"],col=cohortCol,main="PCA",xlab="1st PC", ylab="2nd PC",pch=20)
legend("bottomleft",col=cohortPalette,legend=names(cohortPalette),pch=20)

outliers <- which(pca$rotation[,"PC1"]<.060)
text(x=pca$rotation[outliers,"PC1"],y=pca$rotation[outliers,"PC2"],pos=4,
     labels=gsub("S-161024-","",rownames(pca$rotation)[outliers]))

## let's plot the first and third component
plot(x=pca$rotation[,"PC1"],y=pca$rotation[,"PC3"],col=cohortCol,main="PCA",xlab="1st PC", ylab="3rd PC",pch=20)
legend("topright",col=cohortPalette,legend=names(cohortPalette),pch=20)
text(x=pca$rotation[outliers,"PC1"],y=pca$rotation[outliers,"PC3"],pos=4,
     labels=gsub("S-161024-","",rownames(pca$rotation)[outliers]))

## let's plot the second and third component
plot(x=pca$rotation[,"PC2"],y=pca$rotation[,"PC3"],col=cohortCol,main="PCA",xlab="2nd PC", ylab="3rd PC",pch=20)
legend("topleft",col=cohortPalette,legend=names(cohortPalette),pch=20)
text(x=pca$rotation[outliers,"PC2"],y=pca$rotation[outliers,"PC3"],pos=4,
     labels=gsub("S-161024-","",rownames(pca$rotation)[outliers]))

ord <- order(pData(somaL)$Cohort)
outlier.names <- sampleNames(soma); outlier.names[-outliers] <- ""
outlier.border <- rep("gray",ncol(soma)); outlier.border[outliers] <- "black"
par(mar=c(7, 4, 4, 2))
boxplot(exprs(somaL)[,ord],las=2,pch="-",col=cohortCol[ord],names=outlier.names[ord],
        border=outlier.border[ord],cex.axis=.75)
legend("bottomleft",col=cohortPalette,legend=names(cohortPalette),pch=15,cex=.7)

remove <- sampleNames(soma)[order(pca$rotation[,"PC1"])[1:2]]
soma0 <- somaL[,-match(remove,sampleNames(somaL))]
print(remove)

## add APOE and gen.type

ann <- read.csv(file=file.path(PATH,"data/nastia/data.ss.apoe.csv"))
ann <- data.frame(ann[,c("key.SubjID","APOE","gen.type")],row.names=1)
pdata <- pData(soma0)

pdata <- cbind(pdata,ann[match.nona(rownames(pdata),rownames(ann)),])
APOE3 <- pdata$APOE; APOE3[is.na(APOE3)] <- "e3e4"
pdata$APOE3 <- factor(
    sapply( APOE3, function(Z) if (Z=="e2e3" || Z=="e2e2") "e2" else if (Z=="e3e3") "e3" else if (Z=="e3e4") "e4" else NA),
    levels=c("e2","e3","e4"))
pData(soma0) <- pdata
saveRDS(soma0,file=file.path(PATH,"workdir/somascan/soma0.RDS"))
##soma0 <- readRDS(file.path(PATH,"workdir/somascan/soma0.RDS"))
```

## Analyte Outlier Filtering

Based on the heatmap shown above, and the dominance of very extreme
values, we cap values at +/- 3sd from the (trimmed) mean.

```{r outlier.filtering}
soma1 <- soma0
MN <- apply(exprs(soma1),1,mean,trim=0.025)
SD <- apply(exprs(soma1),1,sd)
exprs(soma1)[exprs(soma1) > MN + 3*SD | exprs(soma1) < MN -3*SD] <- NA
saveRDS(soma1,file.path(PATH,"workdir/somascan/soma1.RDS"))
##soma1 <- readRDS(file.path(PATH,"workdir/somascan/soma1.RDS"))

## let us plot the heatmap again, this time w/ the thresholded values
## ..(and the two outliers removed)

annot1 <- annot[sampleNames(soma1),]
hcRow <- hcopt(as.dist(1-cor(t(exprs(soma1)),use="pairwise.complete.obs")),method="ward.D")
hcCol <- hcopt(dist(t(exprs(soma1))),method="ward.D")
pheatmap(exprs(soma1),
         color=colGradient(c("blue","white","red"),length=15),
         annotation_col = annot1,
         annotation_colors = annotCol,
         cluster_rows=hcRow,
         cluster_cols=hcCol,
         show_rownames = FALSE,
         show_colnames = FALSE,
         scale = "row")
```

## Phenotype Grouping

Next, we collapse some of the groups, so as to have a 3-group phenotype: control, offspring, and centenarians.

```{r diffanal.prep}
soma2 <- soma1
cohort <- as.character(pData(soma2)$Cohort)
cohort[grep("Control",cohort)] <- "Control"
cohort <- gsub("Centenarian Offspring","Offspring",cohort)
cohort <- gsub("Centenarian Singleton","Centenarian",cohort)
cohort <- gsub("Centenarian Sib-Pair","Centenarian",cohort)
pData(soma2)$Cohort <- factor(cohort,levels=c("Control","Offspring","Centenarian"))
saveRDS(soma2,file.path(PATH,"workdir/somascan/soma2.RDS"))
##soma2 <- readRDS(file.path(PATH,"workdir/somascan/soma2.RDS"))
```

```{r longevity.grouping, eval=FALSE, echo=FALSE}
## obsolete, but keeping it in (invisible)
## centenarians grouping
cent <- soma2[,soma2$Cohort=="Centenarian"]
cent$FU <- cent$Age-cent$Age.serum; cent$FU[cent$FU<0] <- 0
cent$age.group <- cut(cent$Age.serum,breaks=c(100,105,110,200),right=FALSE,incluse.lowest=TRUE)
cent$FU.2group <- cut(cent$FU,breaks=c(0,2,Inf),labels=c("short","long"),right=FALSE,include.lowest=TRUE)
table(cent$age.group,cent$FU.2group,useNA="ifany")

## offsprings (+control) grouping
rest <- soma2[,soma2$Cohort!="Centenarian"]
rest$FU <- rest$Age-rest$Age.serum; rest$FU[rest$FU<0] <- 0
rest$age.group <- cut(rest$Age.serum,breaks=c(45,71,100),right=FALSE,incluse.lowest=TRUE)
rest$FU.2group <- cut(rest$FU,breaks=c(0,11,Inf),labels=c("short","long"),right=FALSE,include.lowest=TRUE)
#rest$FU.2group[rest$FU.2group=="[0,11)" & rest$Deceased=="No"] <- NA
table(rest$age.group,rest$FU.2group,useNA="ifany")

soma2$FU.2group <- rep("",ncol(soma2))
pData(soma2)[sampleNames(cent),"FU.2group"] <- as.character(cent$FU.2group)
pData(soma2)[sampleNames(rest),"FU.2group"] <- as.character(rest$FU.2group)
soma2$FU.2group <- factor(soma2$FU.2group,levels=c("short","long") )
saveRDS(soma2,file.path(PATH,"workdir/somascan/soma2.RDS"))
##soma2 <- readRDS(file.path(PATH,"workdir/somascan/soma2.RDS"))
```

